---
title: "Getting Started with lancedb"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with lancedb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The `lancedb` R package provides an embedded vector database client using Rust
bindings. It follows a lazy query pattern similar to `dbplyr`: you build up a
query with familiar dplyr verbs, and nothing executes until you call `collect()`.

## Connecting to a Database

```{r}
library(lancedb)
library(dplyr)

# Connect to a local LanceDB directory
con <- lancedb_connect("/tmp/my_lancedb")
print(con)
```

## Creating a Table

Pass any `data.frame` (or `arrow::Table`) to `lancedb_create_table()`:

```{r}
df <- data.frame(
  id    = 1:1000,
  name  = paste0("item_", 1:1000),
  score = runif(1000),
  category = sample(c("A", "B", "C"), 1000, replace = TRUE)
)

tbl <- lancedb_create_table(con, "products", df)
print(tbl)
```

## Table Scans with dplyr

Use `lancedb_scan()` to begin a lazy query, then chain dplyr verbs:

```{r}
results <- lancedb_scan(tbl) %>%
  filter(score > 0.8, category == "A") %>%
  select(id, name, score) %>%
  slice_head(n = 20) %>%
  collect()

head(results)
```

Note that `filter()`, `select()`, and `slice_head()` do **not** execute
anything — they just append operations to the plan. Only `collect()` triggers
execution.

You can inspect the plan at any time:

```{r}
lancedb_scan(tbl) %>%
  filter(score > 0.5) %>%
  select(name, score) %>%
  show_query()
```

## Vector Similarity Search

For vector search, first create a table with a vector (embedding) column.
LanceDB stores fixed-size lists as vector columns:

```{r}
library(arrow)

# Create data with a vector embedding column
n <- 500
embedding_dim <- 128

# Generate sample data with embeddings
ids <- 1:n
texts <- paste0("document_", ids)
categories <- sample(c("science", "history", "art"), n, replace = TRUE)
embeddings <- lapply(ids, function(i) rnorm(embedding_dim))

vec_df <- data.frame(
  id = ids,
  text = texts,
  category = categories,
  stringsAsFactors = FALSE
)
# Add the embedding column as a list column
vec_df$embedding <- embeddings

vec_tbl <- lancedb_create_table(con, "documents", vec_df)
```

Now search using a query vector:

```{r}
query_vec <- rnorm(embedding_dim)

results <- lancedb_search(vec_tbl, query_vec) %>%
  filter(category == "science") %>%
  select(id, text, category) %>%
  slice_head(n = 10) %>%
  collect()

print(results)
```

Results from `lancedb_search()` include a `_distance` column with the
similarity score (lower = more similar).

## Filter Expressions

You can use R expressions or raw strings:

```{r}
# R expressions (translated to LanceDB SQL)
lancedb_scan(tbl) %>%
  filter(score > 0.5 & category %in% c("A", "B")) %>%
  collect()

# Equivalent string filter
lancedb_scan(tbl) %>%
  filter("score > 0.5 AND category IN ('A', 'B')") %>%
  collect()
```

### Supported R filter syntax

| R Expression       | LanceDB Equivalent     |
|--------------------|------------------------|
| `x == 1`           | `x = 1`                |
| `x != 2`           | `x != 2`               |
| `x > 3`            | `x > 3`                |
| `a & b`            | `a AND b`              |
| `a | b`            | `a OR b`               |
| `!x`               | `NOT (x)`              |
| `x %in% c(1,2,3)` | `x IN (1, 2, 3)`       |
| `is.na(x)`         | `x IS NULL`            |
| `between(x, 1, 5)` | `x BETWEEN 1 AND 5`   |
| `stats$strength`   | `stats.strength`       |

## Arrow Output

To get results as an Arrow Table instead of a data.frame:

```{r}
arrow_result <- lancedb_scan(tbl) %>%
  collect(as = "arrow")

class(arrow_result)
```

## Adding and Deleting Data

```{r}
# Add more data
new_data <- data.frame(id = 1001:1010, name = paste0("new_", 1:10),
                       score = runif(10), category = "D")
lancedb_add(tbl, new_data)

# Delete rows matching a predicate
lancedb_delete(tbl, "category = 'D'")

# Check count
lancedb_count_rows(tbl)
```

## Python Comparison

The R API mirrors the Python LanceDB query builder pattern:

**Python:**
```python
table.search(qvec)
  .where("stats.strength > 3")
  .select(["name", "role", "description"])
  .limit(5)
  .to_pandas()
```

**R:**
```{r}
lancedb_search(tbl, qvec) %>%
  filter(stats$strength > 3) %>%
  select(name, role, description) %>%
  slice_head(n = 5) %>%
  collect()
```
