---
title: "Getting Started with lancedbr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with lancedbr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`lancedbr` provides an R interface to LanceDB, a vector database built on
Apache Arrow. It allows you to create tables, add data, and perform
vector similarity search using a simple, backend-agnostic R API.

The current implementation uses the Python LanceDB client via the
`reticulate` package. The public R API is designed to remain stable as
the backend evolves.

## Installation

At the moment, `lancedbr` is installed from source:

```r
# install.packages("devtools")
devtools::install_github("your-org/lancedbr")
```

The first time you use the package, a Python environment may be created
automatically and the Python `lancedb` package installed.

## Connecting to a database

A LanceDB database is typically stored in a local directory.

```{r}
library(lancedbr)

con <- lancedb_connect("~/mydb")
con
```

This creates (or opens) a database at the given path.

## Creating a table

Tables are created from R data. A common pattern is to store an ID,
metadata columns, and a vector column containing numeric embeddings.

```{r}
data <- data.frame(
  id = 1:3,
  text = c("a", "b", "c"),
  vector = I(list(
    c(0.1, 0.2),
    c(0.2, 0.3),
    c(0.4, 0.5)
  ))
)

tbl <- lancedb_create_table(con, "example", data)
tbl
```

### Vector column requirements

The vector column must be provided as a **list-column of numeric vectors
with a fixed length per row**. All vectors must have the same dimension.

When a table is created from an R `data.frame`, `lancedbr` ingests the
vector column using Apache Arrow with a fixed-size list type. This is
required for vector indexing and similarity search in LanceDB and is
handled automatically by the package.

## Adding data

You can append additional rows to an existing table:

```{r}
tbl_add(tbl, data)
```

The same vector column requirements apply when adding data: vectors must
be numeric and have a consistent length.

## Vector search

Perform a similarity search by providing a numeric query vector:

```{r}
res <- tbl_search(
  tbl,
  query = c(0.1, 0.2),
  k = 2
)

res
```

By default, results are returned as a data frame.

## Returning Arrow results

For larger result sets, you may prefer Arrow output. If the `arrow`
package is installed, set `as = "arrow"`:

```{r, eval = requireNamespace("arrow", quietly = TRUE)}
res_arrow <- tbl_search(
  tbl,
  query = c(0.1, 0.2),
  k = 2,
  as = "arrow"
)

res_arrow
```

Arrow output avoids unnecessary data copies and is recommended for
performance-sensitive workflows.

Note that Apache Arrow is also used internally when ingesting vector
data, even if results are returned as data frames.

## Python configuration

By default, `lancedbr` manages its own Python environment using
`reticulate`. Advanced users can override this behavior with environment
variables:

* `LANCEDBR_PYTHON`: path to a specific Python executable
* `LANCEDBR_CONDA_ENV`: name of a conda environment to use

Example:

```r
Sys.setenv(LANCEDBR_CONDA_ENV = "myenv")
```

These variables should be set **before** loading the package.

## Diagnostics

To inspect the active backend, Python configuration, and installed
Python package versions, use:

```{r}
lancedbr_info()
```

This information is useful when debugging or reporting issues.

## Backend architecture

All user-facing functions in `lancedbr` dispatch to an internal backend
layer. The current backend uses `reticulate`, but the API is designed so
that a future native Rust backend can be introduced without changing
user code.

## Summary

This vignette demonstrated how to:

* connect to a LanceDB database
* create and populate a table with vector data
* perform vector similarity search
* retrieve results as data frames or Arrow tables

For more advanced usage, see the package documentation and source
repository.

