---
title: "Getting Started with lancedb"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with lancedb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The `lancedb` R package provides an embedded vector database client using Rust
bindings. It follows a lazy query pattern similar to `dbplyr`: you build up a
query with familiar dplyr verbs, and nothing executes until you call `collect()`.

## Connecting to a Database

```{r}
library(lancedb)
library(dplyr)

# Connect to a local LanceDB directory
con <- lancedb_connect("/tmp/my_lancedb")
print(con)
```

## Creating a Table

Pass any `data.frame` (or `arrow::Table`) to `lancedb_create_table()`:

```{r}
df <- data.frame(
  id    = 1:1000,
  name  = paste0("item_", 1:1000),
  score = runif(1000),
  category = sample(c("A", "B", "C"), 1000, replace = TRUE)
)

tbl <- lancedb_create_table(con, "products", df)
print(tbl)
```

## Table Scans with dplyr

Use `lancedb_scan()` to begin a lazy query, then chain dplyr verbs:

```{r}
results <- lancedb_scan(tbl) %>%
  filter(score > 0.8, category == "A") %>%
  select(id, name, score) %>%
  slice_head(n = 20) %>%
  collect()

head(results)
```

Note that `filter()`, `select()`, and `slice_head()` do **not** execute
anything — they just append operations to the plan. Only `collect()` triggers
execution.

You can inspect the plan at any time:

```{r}
lancedb_scan(tbl) %>%
  filter(score > 0.5) %>%
  select(name, score) %>%
  show_query()
```

## Vector Similarity Search

For vector search, use `lancedb_search()` with a query vector:
```{r}
# Assuming your table has a vector column
query_vec <- rnorm(128)

results <- lancedb_search(tbl, query_vec) %>%
  filter(category == "B") %>%
  select(id, name, category) %>%
  slice_head(n = 10) %>%
  collect()
```

Results include a `_distance` column with the similarity score.

## Filter Expressions

You can use R expressions or raw strings:

```{r}
# R expressions (translated to LanceDB SQL)
lancedb_scan(tbl) %>%
  filter(score > 0.5 & category %in% c("A", "B")) %>%
  collect()

# Equivalent string filter
lancedb_scan(tbl) %>%
  filter("score > 0.5 AND category IN ('A', 'B')") %>%
  collect()
```

## Arrow Output

To get results as an Arrow Table instead of a data.frame:

```{r}
arrow_result <- lancedb_scan(tbl) %>%
  collect(as = "arrow")

class(arrow_result)
```

## Adding and Deleting Data

```{r}
# Add more data
new_data <- data.frame(id = 1001:1010, name = paste0("new_", 1:10),
                       score = runif(10), category = "D")
lancedb_add(tbl, new_data)

# Delete rows matching a predicate
lancedb_delete(tbl, "category = 'D'")

# Check count
lancedb_count_rows(tbl)
```
